---
title: "p8105_hw2_yl5508"
author: "Yifei LIU"
date: 2023/9/30
output: github_document
---

It's a proj for HW2 about DATA WRANGLING.

```{r setup, echo = FALSE, message = FALSE}
library(tidyverse)
library(readxl)
options(scipen = 999)
options(digits = 2)
```

## Problem 1

### Problem 1.1 DATA CLEANING

```{r, message = FALSE}
pols_data =
  read_csv("./datastore/pols-month.csv") |>
  janitor::clean_names() |>
  separate(mon, into = c("year", "month", "day"), "-") |>
  mutate(month = month.name[as.numeric(month)]) |>
  #the same as
  #mutate(month = case_match(month, case_match(month,"01"~"Jan"), ...)
  #OR using
  #mutate(month = case_when(month=="01"~"Jan"), ...)
  mutate(president = case_when(
    prez_gop != "0" ~ "gop",
    prez_dem != "0" ~ "dem"
  )) |>
  select(-c("day", "prez_gop", "prez_dem")) |>
  select(year, month, president, everything())
head(pols_data, 5)
```

**1.1.1** `pols_data` is from `pols-month.csv`, with `r nrow(pols_data)` obs. and `r ncol(pols_data)` variables.  
Break the variable `mon` into three variables `y`, `m`, `day`, then remove the first two variables.  
Variables are arranged in order of `year`, `month`, `president`, ... .

```{r, message = FALSE}
snp_data =
  read_csv("./datastore/snp.csv") |>
  janitor::clean_names() |>
  separate(date, into = c("month", "day", "year"), "/") |>
  mutate(year = as.numeric(year)) |>
  mutate(year = ifelse(year >= 30 & year <= 99,  1900+year, 2000+year)) |>
  mutate(year = as.character(year)) |>
  mutate(month = month.name[as.numeric(month)]) |>
  select(-day) |>
  select(year, month, everything())
head(snp_data, 5)
```

**1.1.2** `snp_data` is from `snp.csv`, with `r nrow(snp_data)` obs. and `r ncol(snp_data)` variables.  
Use function `ifelse()` to correct year format, then switch it back to character type.  
Turn variable `month` from number into month.name format.  
Variables are arranged in order of `year`, `month`, `close`.

```{r, message = FALSE}
unemployment_data =
  read_csv("./datastore/unemployment.csv") |>
  janitor::clean_names() |>
  pivot_longer(
    jan:dec,
    names_to = "month",
    values_to = "unemployment_rate"
  ) |>
  mutate(month = month.name[match(str_to_title(month), month.abb)]) |>
  select(year, month, everything())
head(unemployment_data, 5)
```

**1.1.3** `unemployment_data` is from `unemployment.csv`, with `r nrow(unemployment_data)` obs. and `r ncol(unemployment_data)` variables.   
Reorganize the dataset by creating two new variables `month` and `unemployment`, using function `pivot_longer()`.  
Variables are arranged in order of `year`, `month`, `unemployment_rate`.

### Problem 1.2 JOINING DATASETS

Create variable `year_month` combining variable `year` and `month`, then merge the data by variable `year_month`.

```{r, message = FALSE}
pols_join =
  pols_data |>
  mutate(year_month = paste(year, month, sep = "-")) |>
  select(year_month, everything())

snp_join =
  snp_data |>
  mutate(year_month = paste(year, month, sep = "-")) |>
  select(year_month, everything())

unemployment_join =
  unemployment_data |>
  mutate(year_month = paste(year, month, sep = "-")) |>
  select(year_month, everything())

pols_snp = left_join(pols_join, snp_join, by = "year_month") |>
  select(-year.y, -month.y) |>
  select(year_month, year = year.x, month = month.x, president, close, everything()) |>
  arrange(year)
head(pols_snp, 5)

pols_snp_unemployment = left_join(pols_snp, unemployment_join, by = "year_month") |>
  select(-year.y, -month.y) |>
  select(year_month, year = year.x, month = month.x, president, close, unemployment_rate, everything()) |>
  arrange(year)
head(pols_snp_unemployment, 5)
```

### Problem 1.3 DATA REVIEW

We can use `skim()` to grab some key characters of the dataset.

```{r, message = FALSE}
skimr::skim(pols_snp_unemployment)
```

(a) `pols_data` contains `r nrow(pols_data)` obs. of `r ncol(pols_data)` variables. The key variables are `year`, `month`, `president`.  
(b) `snp_data` contains `r nrow(snp_data)` obs. of `r ncol(snp_data)` variables. The key variables are `year`, `month`, `close`.  
(c) `unemployment_data` contains `r nrow(unemployment_data)` obs. of `r ncol(unemployment_data)` variables. The key variables are `year`, `month`, `unemployment_rate`.

## Problem 2

### Problem 2.1 DATA CLEANING

```{r, message = FALSE}
mrt_data =
  read_excel("./datastore/202309 Trash Wheel Collection Data.xlsx", sheet = "Mr. Trash Wheel", range = "A2:N586") |>
  janitor::clean_names() |>
  separate(date, into=c("y", "m", "day"), sep = "-") |>
  select(-y, -m) |>
  mutate(year = as.character(year)) |>
  mutate(homes_powered = weight_tons*500/30) |>
  mutate(sourcename = "Mr. Trash Wheel") |>
  select(sourcename, dumpster, year, month, day, homes_powered, weight_tons, everything())
head(mrt_data, 5)
```

**2.1.1** `mrt_data` is from sheet `Mr. Trash Wheel`, with `r nrow(mrt_data)` obs. and `r ncol(mrt_data)` variables.  
Break the variable `date` into three variables `y`, `m`, `day`, then remove the first two variables and get variable `day`.  
Calculate the `homes_powered` using data in `weight_tons`, according to the given formula.  
Create a label(column `sourcename`) for this dataset.  
Variables are arranged in order of `sourcename`, `dumpster`, `year`, `month`, `day`, `homes_powered`, `weight_tons` ... .

```{r, message = FALSE}
pro_data =
  read_excel("./datastore/202309 Trash Wheel Collection Data.xlsx", sheet = "Professor Trash Wheel", range = "A2:M108") |>
  janitor::clean_names() |>
  separate(date, into=c("y", "m", "day"), sep = "-") |>
  select(-y, -m) |>
  mutate(year = as.character(year)) |>
  mutate(homes_powered = weight_tons*500/30) |>
  mutate(sourcename = "Professor Trash Wheel") |>
  select(sourcename, dumpster, year, month, day, homes_powered, weight_tons, everything())
head(pro_data, 5)
```

**2.1.2** `pro_data` is from sheet `Professor Trash Wheel`, with `r nrow(pro_data)` obs. and `r ncol(pro_data)` variables.  
Break the variable `date` into three variables `y`, `m`, `day`, then remove the first two variables and get variable `day`.  
Calculate the `homes_powered` using data in `weight_tons`, according to the given formula.  
Create a label(column `sourcename`) for this dataset.  
Variables are arranged in order of `sourcename`, `dumpster`, `year`, `month`, `day`, `homes_powered`, `weight_tons` ... .

```{r, message = FALSE}
gwy_data =
  read_excel("./datastore/202309 Trash Wheel Collection Data.xlsx", sheet = "Gwynnda Trash Wheel", range = "A2:L157") |>
  janitor::clean_names() |>
  separate(date, into=c("y", "m", "day"), sep = "-") |>
  select(-y, -m) |>
  mutate(year = as.character(year)) |>
  mutate(homes_powered = weight_tons*500/30) |>
  mutate(sourcename = "Gwynnda") |>
  select(sourcename, dumpster, year, month, day, homes_powered, weight_tons, everything())
head(gwy_data, 5)
```

**2.1.3** `gwy_data` is from sheet `Gwynnda Trash Wheel`, with `r nrow(gwy_data)` obs. and `r ncol(gwy_data)` variables.   
Break the variable `date` into three variables `y`, `m`, `day`, then remove the first two variables and get variable `day`.  
Calculate the `homes_powered` using data in `weight_tons`, according to the given formula.  
Create a label(column `sourcename`) for this dataset.  
Variables are arranged in order of `sourcename`, `dumpster`, `year`, `month`, `day`, `homes_powered`, `weight_tons` ... .  

**2.1.4** Then merge all three datasets together.

```{r, message = FALSE}
total_trash_data = 
  bind_rows(mrt_data, pro_data, gwy_data) |>
  arrange(year, match(month, month.name), day)
head(total_trash_data, 5)
```

### Problem 2.2 DATA REVIEW

We can use `skim()` to grab some key characters of the dataset. 

```{r, message = FALSE}
skimr::skim(total_trash_data)
```

(a) The total number of observations in result dataset is `r nrow(total_trash_data)`. Variables number in result dataset is `r ncol(total_trash_data)`.
(b) The variables included in dataset are `sourcename`, `dumpster`, `year`, `month`, `day`, `homes_powered`, `weight_tons`, `volume_cubic_yards`, `plastic_bottles`, `polystyrene`, `cigarette_butts`, `plastic_bags`, `wrappers`, ... . Key variables `year`, `month`, `day` indicate the record time, `weight_tons` indicates the amount of trash collected, `homes_powered` indicates number of homes being powered by using trash recycle tech, and `sourcename` indicates from which sheet we get the corresponding data.
(c) Total weight collected by Professor Trash Wheel is `r total_trash_data |> filter(sourcename == "Professor Trash Wheel") |> summarise(sum(weight_tons)) |> pull()` (or using another formula `r sum(total_trash_data$weight_tons[total_trash_data$sourcename == "Professor Trash Wheel"])`).
(d) Total number of cigarette butts collected by Gwynnda in July of 2021 is `r total_trash_data |> filter(sourcename == "Gwynnda" & year == "2021" & month == "July") |> summarise(sum(cigarette_butts)) |> pull()`.

## Problem 3

### Problem 3.1 DATA CLEANING

For dataset `mci_baseline.csv` and `mci_amyloid.csv`, we notice that there're two rows delineating info of the data. During cleaning process, we shall skip them in order to grab the raw data we want.

```{r}
bas_data_clean =
  read_csv("./datastore/MCI_baseline.csv", skip = 1) |>
  janitor::clean_names() |>
  mutate(sex = case_when(
    sex == 1 ~ "Male",
    sex == 0 ~ "Female"
  )) |>
  mutate(apoe4 = case_when(
    apoe4 == 1 ~ "carrier",
    apoe4 == 0 ~ "non-carrier"
  )) |>
  mutate(age_at_onset = ifelse(age_at_onset == '.', NA, age_at_onset)) |>
  filter(current_age < age_at_onset | is.na(age_at_onset)) |>
  mutate(age_at_onset = as.numeric(age_at_onset))
head(bas_data_clean, 5)

bas_data_dropna =
  bas_data_clean |>
  drop_na(age_at_onset)
head(bas_data_dropna)

bas_data =
  read_csv("./datastore/MCI_baseline.csv", skip = 1) |>
  janitor::clean_names() |>
  mutate(sex = case_when(
    sex == 1 ~ "Male",
    sex == 0 ~ "Female"
  )) |>
  mutate(apoe4 = case_when(
    apoe4 == 1 ~ "carrier",
    apoe4 == 0 ~ "non-carrier"
  )) |>
  mutate(age_at_onset = ifelse(age_at_onset == '.', NA, age_at_onset))
head(bas_data, 5)

amy_data =
  read_csv("./datastore/mci_amyloid.csv", skip = 1) |>
  janitor::clean_names() |>
  select(id = study_id, year_0 = baseline, year_2 = time_2, year_4 = time_4, year_6 = time_6, year_8 = time_8) |>
  pivot_longer(
    year_0:year_8,
    names_to = "time",
    values_to = "ratio"
  )
head(amy_data, 5)
```

### Problem 3.2 DATA REVIEW

(a) There're `r nrow(bas_data)` participants recruited in the research, in which `r nrow(bas_data_clean)` participants meet the criteria. From the data, we notice `r nrow(bas_data_dropna)` participants develop MCI (deem NA as healthy condition).
(b) The average baseline age is `r mean(pull(bas_data_clean, current_age))`. For participants developing MCI, the average baseline age is `r mean(pull(bas_data_dropna, current_age))`.
(c) The proportion of women are APOE4 carriers is `r (count(bas_data_clean |> filter(sex=="Female" & apoe4=="carrier")) |> pull()) / (count(bas_data_clean |> filter(sex=="Female")) |> pull())`.
(d) 

```{r}
#manipulate in dataset
baseline_only = anti_join(bas_data_clean, amy_data, by = "id")
amyloid_only = anti_join(amy_data, bas_data_clean, by = "id")
```

```{r}
#grab diff data
only_baseline = setdiff(bas_data_clean$id, amy_data$id)
only_amyloid = setdiff(amy_data$id, bas_data_clean$id)
```

After processing, `r nrow(baseline_only)` participants only appear in baseline dataset. IDs of these participants are shown below: 
`r only_baseline`.  

`r nrow(amyloid_only)/5` participants only appear in baseline dataset. IDs of these participants are shown below:
`r only_amyloid`. 

(e) 
```{r}
#combine two datasets into one, with only participants appearing in both datasets
bas_amy = inner_join(bas_data_clean, amy_data, by = "id")
```
There're `r length(unique(pull(bas_amy, id)))` participants appear in both dataset. Because ratio data of each participant is posted for 5 times (in different time intervals), ID of every participant has been shown 5 times. `r ncol(bas_amy)` variables are listed in combined dataset, included `id`, `current_age`, `sex`, `education`, `apoe4`, `age_at_onset`, `time`, `ratio`.  

(f) The combined data has been saved as `combined_bas_amy.csv`.

```{r}
write.csv(bas_amy, file = "./datastore/combined_bas_amy.csv")
```

